One-loop PDF of the average Ising magnetization in 4 dimensions.

### 2020-03-20 paper/paper.tex
Paper checkpoint

Thus, because of the delta function in (\ref{eq:pdf_definition}), the component
$\phi_{0}$ becomes an external field, whereas the other components $\phi_{n}$
with are still part of the functional integral. Separating the zero momentum
mode from the other modes in the action yields:
\begin{equation}
\begin{split}
    S(\bar{\phi}, \phi) =& 
    V \left(\frac{1}{2} r_0 \bar{\phi}^2 + \frac{u_0}{24} \bar{\phi}^4\right) +
    \frac{u_0}{4V} \bar{\phi}^2 \sum_{n} \phi_{n} \phi_{-n} + \\
    &\frac{u_0}{6V^2} \bar{\phi} 
        \sum_{n_1, n_2} \phi_{n_1} \phi_{n_2} \phi_{-n_1 -n_2} + \\
    &\frac{u_0}{24V^3} \sum_{n_1, n_2, n_3} \phi_{n_1} \phi_{n_2} 
        \phi_{n_3} \phi_{-n_1 -n_2 -n_3}\,.
\end{split}
\end{equation}

We also introduce renormalized couplings
\begin{equation}
    u_0 = u(1 + u\delta u)\,;\quad r_0 = r(1 + u \delta r)\,,
\end{equation}
and we obtain the following edges and vertices in the diagrammatic expansion:
\begin{displaymath}
\begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
    \draw[dashed, thick] (-1, -1) -- (0, 0) -- (-1, 1);
    \draw[thick] (+1, -1) -- (0, 0) -- (+1, 1);
    \node[above right] at (1, 1) {$n_1$};
    \node[below right] at (1, -1) {$n_2$};
\end{tikzpicture}
    -\frac{u\bar{\phi}^2}{4V}\delta_{n_1 + n_2}
    \quad\quad
\begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
    \draw[dashed, thick] (-1, -1) -- (0, 0);
    \draw[thick] (0, 0) -- (-1, 1);
    \draw[thick] (+1, -1) -- (0, 0) -- (+1, 1);
    \node[above left] at (-1, 1) {$n_1$};
    \node[above right] at (1, 1) {$n_2$};
    \node[below right] at (1, -1) {$n_3$};
\end{tikzpicture}
    -\frac{u\bar{\phi}}{6V^2}\delta_{n_1 + n_2 + n_3}
\end{displaymath}
\begin{displaymath}
\begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
    \draw[thick] (-1, -1) -- (1, 1);
    \draw[thick] (-1, 1) -- ( 1, -1);
    \node[above left] at (-1, 1) {$n_1$};
    \node[above right] at (1, 1) {$n_2$};
    \node[below right] at (1, -1) {$n_3$};
    \node[below left] at (-1, -1) {$n_4$};
\end{tikzpicture}
    -\frac{u}{24V^3}\delta_{n_1 + n_2 + n_3 + n_4}
\quad\quad
\begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
    \draw[thick] (-1, 0) -- (1, 0);
    \node[left] at (-1, 0) {$n_1$};
    \node[right] at (1, 0) {$n_2$};
\end{tikzpicture}
    \quad
    \frac{V}{n_1^2R^{-2} + r}\delta_{n_1 + n_2}
\end{displaymath}
plus additional vertices associated with the counterterms $\delta r$, $\delta
u$ which we do not list for brevity.

The logarithm of the probability distribution $p(\bar{\phi}$ is the sum of all connected diagrams:
\begin{displaymath}
\begin{split}
    \log p(\bar{\phi}) = &\,
    \mathrm{const} -V \left(\frac{1}{2} r \bar{\phi}^2 + 
    \frac{u}{24} \bar{\phi}^4\right)\\
&+\ 
\begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
    \draw[thick, dashed] (-1, -0.5) -- (0, 0) -- (1, -0.5);
    \draw[thick] (0, 0) .. controls (-1.5, 1.3) and (1.5, 1.3) .. (0, 0);
\end{tikzpicture}
\ + \ 
\begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
    \draw[thick, dashed] (-1.75, 0) -- (-0.75, 0);
    \draw[thick] (-0.75, 0) -- (0.75, 0);
    \draw[thick, dashed] (1.75, 0) -- (0.75, 0);
    \draw[thick] (0, 0) circle (0.75cm);
\end{tikzpicture}
\ + \ 
\begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
    \draw[thick, dashed] (-1, -0.5) -- (0, 0) -- (1, -0.5);
    \draw[thick] (0, 0) .. controls (0.6, 0.5).. (0, 1) .. 
    controls (-1.5, 2.3) and (1.5, 2.3) .. (0, 1) .. 
    controls (-0.6, 0.5) .. (0, 0) ;
\end{tikzpicture}
 \ + \ \cdots \\
&+\ 
\begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
    \draw[thick, dashed] (-1.75, -0.75) -- (-1, 0) -- (-1.75, 0.75);
    \draw[thick, dashed] (1.75, -0.75) -- (1, 0) -- (1.75, 0.75);
    \draw[thick] (-1, 0) .. controls (-0.2, 0.8) and (0.2, 0.8) .. (1, 0) .. 
    controls (0.2, -0.8) and (-0.2, -0.8) .. (-1, 0);
\end{tikzpicture}
\ + \ 
\begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
    \draw[thick, dashed] (-1.5, 0.75) -- (-0.75, 0) -- (-1.5, -0.75) ;
    \draw[thick] (0, 0) circle (0.75cm);
    \draw[thick, dashed] (0.53, 0.53) -- (1.25, 1.25);
    \draw[thick, dashed] (0.53, -0.53) -- (1.25, -1.25);
    \draw[thick] (0.53, -0.53) .. controls (0.2, -0.2) 
    and (0.2, 0.2) .. (0.53, 0.53);
\end{tikzpicture}
\ + \ 
\begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
    \draw[thick, dashed] (-1.75, -0.75) -- (-1, 0) -- (-1.75, 0.75);
    \draw[thick, dashed] (1.75, -0.75) -- (1, 0) -- (1.75, 0.75);
    \draw[thick] (-1, 0) .. controls (-0.2, 0.8) and (0.2, 0.8) .. (1, 0) .. 
    controls (0.2, -0.8) and (-0.2, -0.8) .. (-1, 0);
\end{tikzpicture}
\end{split}
\end{displaymath}

>>>>

Thus, because of the delta function in (\ref{eq:pdf_definition}), the component
$\phi_{0}$ becomes an external field, whereas the other components $\phi_{n}$
with are still part of the functional integral. Separating the zero momentum
mode from the other modes in the action yields:
\begin{equation}
\begin{split}
    S(\bar{\phi}, \phi) =& 
    V \left(\frac{1}{2} r_0 \bar{\phi}^2 + \frac{u_0}{24} \bar{\phi}^4\right) +
    \frac{u_0}{4V} \bar{\phi}^2 \sum_{n} \phi_{n} \phi_{-n} + \\
    &\frac{u_0}{6V^2} \bar{\phi} 
        \sum_{n_1, n_2} \phi_{n_1} \phi_{n_2} \phi_{-n_1 -n_2} + \\
    &\frac{u_0}{24V^3} \sum_{n_1, n_2, n_3} \phi_{n_1} \phi_{n_2} 
        \phi_{n_3} \phi_{-n_1 -n_2 -n_3}\,,
\end{split}
\end{equation}
where all summations now exclude $n = (0, 0, 0, 0)$.

We also introduce renormalized couplings
\begin{equation}
    u_0 = u(1 + u\delta u)\,;\quad r_0 = r(1 + u \delta r)\,,
\end{equation}
and we obtain the following edges and vertices in the diagrammatic expansion:
\begin{displaymath}
\begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
    \draw[densely dashed, thick] (-1, -1) -- (0, 0) -- (-1, 1);
    \draw[thick] (+1, -1) -- (0, 0) -- (+1, 1);
    \node[above right] at (1, 1) {$n_1$};
    \node[below right] at (1, -1) {$n_2$};
\end{tikzpicture}
    -\frac{u\bar{\phi}^2}{4V}\delta_{n_1 + n_2}
    \quad\quad
\begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
    \draw[densely dashed, thick] (-1, -1) -- (0, 0);
    \draw[thick] (0, 0) -- (-1, 1);
    \draw[thick] (+1, -1) -- (0, 0) -- (+1, 1);
    \node[above left] at (-1, 1) {$n_1$};
    \node[above right] at (1, 1) {$n_2$};
    \node[below right] at (1, -1) {$n_3$};
\end{tikzpicture}
    -\frac{u\bar{\phi}}{6V^2}\delta_{n_1 + n_2 + n_3}
\end{displaymath}
\begin{displaymath}
\begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
    \draw[thick] (-1, -1) -- (1, 1);
    \draw[thick] (-1, 1) -- ( 1, -1);
    \node[above left] at (-1, 1) {$n_1$};
    \node[above right] at (1, 1) {$n_2$};
    \node[below right] at (1, -1) {$n_3$};
    \node[below left] at (-1, -1) {$n_4$};
\end{tikzpicture}
    -\frac{u}{24V^3}\delta_{n_1 + n_2 + n_3 + n_4}
\quad\quad
\begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
    \draw[thick] (-1, 0) -- (1, 0);
    \node[left] at (-1, 0) {$n_1$};
    \node[right] at (1, 0) {$n_2$};
\end{tikzpicture}
    \quad
    \frac{V}{n_1^2R^{-2} + r}\delta_{n_1 + n_2}
\end{displaymath}
plus additional vertices associated with the counterterms $\delta r$, $\delta
u$ which we do not list for brevity.

The logarithm of the probability distribution of $\bar{\phi}$ is the sum of all
connected diagrams:
\begin{displaymath}
\begin{split}
    \log p(\bar{\phi}) = &\,
    \mathrm{const} -V \left(\frac{1}{2} r \bar{\phi}^2 + 
    \frac{u}{24} \bar{\phi}^4\right)\\
&+\ 
\begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
    \draw[thick, densely dashed] (-1.75, -0.75) -- (-1, 0) -- (-1.75, 0.75);
    \draw[thick] (-1, 0) .. controls (-0.2, 0.8) and (0.5, 0.8) .. (0.5, 0) .. 
    controls (0.5, -0.8) and (-0.2, -0.8) .. (-1, 0);
\end{tikzpicture}
\ + \ 
\begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
    \draw[thick, densely dashed] (-1.75, 0) -- (-0.75, 0);
    \draw[thick] (-0.75, 0) -- (0.75, 0);
    \draw[thick, densely dashed] (1.75, 0) -- (0.75, 0);
    \draw[thick] (0, 0) circle (0.75cm);
\end{tikzpicture}
\ + \ 
\begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
    \draw[thick, densely dashed] (-2.75, -0.75) -- (-2, 0) -- (-2.75, 0.75);
    \draw[thick] (-2, 0) .. controls (-1.2, 0.8) and (-0.8, 0.8) .. (0, 0) .. 
    controls (-0.8, -0.8) and (-1.2, -0.8) .. (-2, 0);
    \draw[thick] (1.6, 0) .. controls (1.6, 0.8) and (0.8, 0.8) .. (0, 0) .. 
    controls (0.8, -0.8) and (1.6, -0.8) .. (1.6, 0);
\end{tikzpicture}
 \ + \ \cdots \\
&+\ 
\begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
    \draw[thick, densely dashed] (-1.75, -0.75) -- (-1, 0) -- (-1.75, 0.75);
    \draw[thick, densely dashed] (1.75, -0.75) -- (1, 0) -- (1.75, 0.75);
    \draw[thick] (-1, 0) .. controls (-0.2, 0.8) and (0.2, 0.8) .. (1, 0) .. 
    controls (0.2, -0.8) and (-0.2, -0.8) .. (-1, 0);
\end{tikzpicture}
\ + \ 
\begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
    \draw[thick, densely dashed] (-1.5, 0.75) -- (-0.75, 0) -- (-1.5, -0.75) ;
    \draw[thick] (0, 0) circle (0.75cm);
    \draw[thick, densely dashed] (0.53, 0.53) -- (1.25, 1.25);
    \draw[thick, densely dashed] (0.53, -0.53) -- (1.25, -1.25);
    \draw[thick] (0.53, -0.53) .. controls (0.2, -0.2) 
    and (0.2, 0.2) .. (0.53, 0.53);
\end{tikzpicture}
\ + \ 
\begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
    \draw[thick, densely dashed] (-2.75, -0.75) -- (-2, 0) -- (-2.75, 0.75);
    \draw[thick, densely dashed] (2.75, -0.75) -- (2, 0) -- (2.75, 0.75);
    \draw[thick] (-2, 0) .. controls (-1.2, 0.8) and (-0.8, 0.8) .. (0, 0) .. 
    controls (-0.8, -0.8) and (-1.2, -0.8) .. (-2, 0);
    \draw[thick] (2, 0) .. controls (1.2, 0.8) and (0.8, 0.8) .. (0, 0) .. 
    controls (0.8, -0.8) and (1.2, -0.8) .. (2, 0);
\end{tikzpicture}
 \ + \ \cdots \\
& + \ 
\begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
    \draw[thick, densely dashed] (-1.5, 0.75) -- (-0.75, 0) -- (-1.5, -0.75) ;
    \draw[thick] (0, 0) circle (0.75cm);
    \draw[thick, densely dashed] (1.40, -0.92) -- (0.375, -0.65) -- (0.1, -1.67);
    \draw[thick, densely dashed] (1.40, 0.92) -- (0.375, 0.65) -- (0.1, 1.67);
\end{tikzpicture}
 \ + \ \cdots \,,
\end{split}
\end{displaymath}
where again we left out all diagrams involving the counterterms for brevity.

Retaining only one-loop diagrams, and employing a heath kernel regulator we have:
\begin{equation}
\begin{split}
    -\log p(\bar{\phi}) = \mathrm{const} + V \Bigg( 
    &\frac{1}{2} \bar{\phi}^2\left( r + u \left(\delta r + \frac{1}{2} I_1\right) + O(u^2)\right) + \\
    &\frac{u}{24} \bar{\phi}^4 \left(1 + u \left(\delta u - \frac{3}{2} I_2\right) + O(u^2)\right) + \\
    &\frac{u^3}{48}\bar{\phi}^6\left(I_3 + O(u)\right) + O(u^4 \bar{\phi}^8) \Bigg)\,,
\end{split}
\end{equation}
where
\begin{equation}
    I_k = \frac{1}{V} \sum_{n\in\mathbb{Z}^4\setminus\{0\}} 
    \frac{e^{-s(r + n^2 R^{-2})}}{\left(r + n^2 R^{-2}\right)^k} = 
    \frac{1}{(k - 1)!}\left(-\frac{\partial}{\partial r} - s\right)^k I_1
\end{equation}

The summation $I_1$ can be simplified substantially using the following trick:
\begin{equation}
\begin{split}
    I_1 &= \frac{1}{V} \int_{s}^{\infty} \dd t \sum_{n\in\mathbb{Z}^4\setminus\{0\}} 
    e^{-t(r + n^2 R^{-2})}\\
    & = \frac{1}{V} \int_{s}^{\infty} \dd t\ e^{-t r}
        \left(\left(\sum_{m = -\infty}^{\infty} e^{-t m^2 R^{-2}}\right)^4 - 1\right) \\
    & = \frac{1}{V} \int_{s}^{\infty} \dd t\ e^{-t r}
        \left(\theta\!\left(t R^{-2}\right)^4 - 1\right)\,,
\end{split}
\end{equation}
where:\footnote{The function $\theta$ is in fact a Jacobi theta function, but we
won't need any of its special properties beyond its asymptotic behavior, which
can be easily obtained from the definition.}
\begin{equation}
    \theta(z) \equiv \sum_{m = -\infty}^{\infty} e^{-m^2 z}\,.
\end{equation}

We now extract the divergent part of $I_1$ by subtracting under the integral a
function that has the same asymptotic behavior as the integrand for $t \to 0$,
but whose integral can be computed in closed form. In order to do that, we need
the asymptotic behavior of $\theta(z)$ for $z\to0$, which can be obtained from
its definition using the Euler-Mclaurin formula:
\begin{equation}
    \theta(z) \sim \sqrt{\frac{\pi}{z}} + o(z^k) \quad\mathrm{for}\, z\to0\,.
\end{equation}

Thus we have:
\begin{equation}
\begin{split}
    I_1 &= \frac{1}{16\pi^2}\left(\int_{s}^{\infty}\dd t\, e^{-t \mu^2} t^{-2}\left(1 + t (\mu^2 - r)\right)
    +\frac{1}{R^2} f(r R^2, \mu R)\right)\\
    &= \frac{1}{16\pi^2}\left(\frac{1}{s} + r\left(\log s\mu^2 + \gamma_{E}\right) - 
    \mu^2 +\frac{1}{R^2} f(r R^2, \mu R)\right)\,,
\end{split}
\end{equation}
where $\mu$ is a renormalization scale that can be chosen at will, and:
\begin{equation}
    f(rR^2,\mu R) = \int_{0}^{\infty} \dd z\ e^{-z r R^2} \left(\frac{\theta(z)^4 - 1}{\pi^2} - 
    \frac{1 + z\left(\mu^2 - r\right)R^2}{t^2}e^{-z\left(\mu^2 - r\right)R^2}\right).
\end{equation}
